---
title: "Single Trial EEG Data Analysis and BVAtoR"
output: rmarkdown::html_vignette
author: "Sarah von Grebmer"
date: "25 April 2020"
vignette: >
  %\VignetteIndexEntry{Single Trial EEG Data Analysis and BVAtoR}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(magrittr)
```


```{r setup}
library(BVAtoR)
library(ggplot2)
```

The BVAtoR package takes segmented output from BVA (.dat files, .vmrk files and .vhdr files) and transforms them into a single file with all voltage values, participants, conditions, and other variables specified by the experimenter for further single trial statistical analysis. This vignette describes the installation and use of the package in order to obtain a (large) dataframe containing EEG data from BVA output. 

# BVAtoR Pre-Requisites

This package works excusively with BVA (V2.2) export files, i.e. pre-processed, segmented and baseline-corrected data (not averaged data). An important step during segmentation in BVA is to assign the resulting node a meaningful name, ideally containing information about the participant (i.e. participant number) and the condition, separated by underscores (or any other separator). 
Upon termination of the classical pre-processing procedure (e.g. re-referencing, filtering, ocular correction, artefact rejection, segmentation, baseline correction) in BVA, the data is ready for export. Note that each condition for each participant will have to be exported separately. See below for instructions on how to export data from BVA. 

## Exporting Data from BVA

1. In BVA, select the node from which you want to export data, namely the baseline corrected node containing information about single trials.

2. In BVA, click EXPORT/GENERIC DATA/ tick WRITE HEADER FILE and WRITE MARKER FILE/ select TEXT FORMAT for the header and marker file/include participant number and condition in the basename (CRUCIAL!)/ click NEXT

3. Select TEXT FORMAT as data file format/MULTIPLEXED as data orientation/ LINE DELIMITERS as PC format/ click NEXT

4. Tick ADD CHANNEL NAMES TO THE DATAFILE/ Tick OVERWRITE DEFAULT DECIMAL SYMBOL (,) and replace to DOT/CUSTOM PRECISION TO SIX OR MORE/ change value layout to SEPARATE VALUES BY CUSTOM SEPARATOR/ set separator to semi-colon (;)/ click NEXT

5. Untick EXPORT ALL CHANNELS/ remove VEOG, HEOG AND ALL EYE CHANNELS from selected channels / click FINISH

This will create three separate export files in your EXPORT folder. Make a copy of these files and copy all files to a separate folder, which will be the working directory of this current pipeline. 

* dat file: contains the voltage values for each channel
* vmrk file: markers text file that contains the order of appearance of the markers and their timestamp in samples
* vhdr file: header file which contains metadata such as sampling frequency and the link between the two other export files

Note: Example File Name
To illustrate the naming of files, the file subj100_comp_BaselineCorrection_cogn_incongr_nonviol.dat contains the following information:

subj100: subject number
comp: Name of Task
Baseline_Correction: Node of export in BVA (i.e. exported after baseline correction before averaging)
cogn: Level 1 of Condition 1
congr: Level 1 of Condition 2
non-viol: Level 1 of Condition 3

# BVAtoR Package

The BVAtoR package was specifically created to combine the voltage values in the .dat file with the information about the markers in the .vmrk file to generate a large datafile that contains the variables "marker" and "voltage", among other variables, such as condition etc. 

```{r installBVAtoR, eval=FALSE}
# Load package
library(BVAtoR)
```

# BVA_to_R Function: Parameters

BVAtoR is based on a number of helper functions which tackle the formatting of the export files and the extraction of crucial information, such as the sampling frequency or the marker names. The helper functions run in the background of the core function of this package, the BVA_to_R() function. Please see below for a more detailed description of the helper functions. 
The core function takes a number of parametersthat should be known to the researcher prior to starting this analysis: 

* path: path to the data folder with all exort files, e.g. "template/path/to/all/files/" . Specified outside the main function.

* full_window_bounds: The total length of the exported segment in ms (e.g. c(-200, 1200). Specified within the main function.

* ERP_list: List of ERP components of interest with their associated lower and upper time limits. These should correspond to the time windows of the ERP component sand cover the entire segment. See below for an example. Specified outside the main function.

Note: Ensure to cover the entire length of the segment. Prior to setting ERP time limits, the researcher must be aware of the number of time samples taken during the duration of a segment and the sampling rate. The sampling frequency (SF) in Hz is needed to calculate this time interval, and can be retrieved from the .vhdr file. 

If we assume a sampling frequency (SF) of 500Hz the time inbetween voltage samples (in milliseconds ms) is calculated as follows: 

1000 ms/SF = Time_Increment
1000 ms/500 = Time_Increment
1000 ms/500 = 2 ms

This reads as: at a sampling frequency of 500 Hz, voltage is measured every 2 ms. Now we need to calculate the number of voltage samples per segment

To calculate the number of voltage samples taken during a segment of (for example) 1400 ms: 

Total_segment_Length in ms/ Time_Increment = Number_of_Samples
1400 ms/2 ms = 700 voltage samples

In conclusion, the time bounds for each ERP component cannot be odd numbers. See below for an example. Further parameters to be specified are:  

* var_labs : variables to inlcude as columns in the final datafile. These correspond to the information given in the exported data file names, i.e. participant, condition etc. Specified within the main function.

* sep: specifies the separator with which the variables within the exported data file names are separated by. Specified within the main function.

* filename_returned: the name of the file to be returned in csv format, e.g. "Output.csv". Contains columns for subject, marker, condition (if applicable), timestamp, ERP component (if applicable), node of segmentation, task (if applicable) etc. for further single trial statisitical analysis in R. Specified within the main function.

## Define BVA_to_R Parameters

```{r defineparameters}
# Help to main function
#?BVA_to_R

# Specify path to export files
path <- "../inst/extdata/"

# Specify list of ERPs (example), ensure to cover entire window length and that they do not overlap
ERPs <- list(
"baseline" = c(-200, 48), # pre-stimulus
"N100" = c(48, 198),
"P200" = c(200, 348),
"N400" = c(350, 498),
"P600" = c(500, 898), # longer P600 window for late learners (example-specific)
"PostP600" = c(900, 1198) # end of segment
)
```

# BVA_to_R Function: Useage

```{r corefunction}
# Call main function and generate final datafile (obj)
obj <- BVA_to_R(path = path, 
                sep = "_", 
                ERP_list = ERPs, 
                var_labs = c(
                  "Subject", "Comp", "Export_Node", 
                  "Cognate_Status", "Congruency_Type", "Violation_Type"
                ),
                full_window_bounds = c(-200, 1198)) 

print(obj)
```

## Other functions

### ERP_to_times

This function converts ERP List to dataframe with ERP components and time vectors.

```{r ERPsToTimes}
times <- seq(-200, 1198)

ERP_to_times(ERPs, times) %>% 
  data.table::data.table()
```

### get_markers

Formats the vmrk files to extract information about markers

```{r formatvmrk}
get_markers("../inst/extdata/kons104_comp_BaselineCorrection_noncogn_incongr_nonviol.vmrk") %>% 
  data.table::data.table()
```

### one_vhdr

BVA_to_R function but for a single vhdr file

```{r BVAtoRsinglePPT}
file <- "../inst/extdata/kons104_comp_BaselineCorrection_noncogn_incongr_nonviol.vhdr"
path <- "../inst/extdata/"

one_vhdr(
  path = path,
  filename = file,
  sep = "_", 
  ERP_list = ERPs, 
  var_labs = c(
     "Subject", "Comp", "Export_Node",
     "Cognate_Status", "Congruency_Type", "Violation_Type"             
  ),
  full_window_bounds = c(-200, 1198)
)
```

# Session Information

```{r session}
sessionInfo()
```